The evolution of gene expression in the term placenta of viviparous mammals

### 2016-03-24 manuscript/manuscript.Rnw
remove ouch discussion

All 1:1 orthologs present among \opossum, \textit{Pan troglodytes},
\cow, \dog, \elephant, \mouse, and \human were determined using
Ensembl v.80 gene trees via Biomart
\cite{Kasprzyk2011:BioMartdrivingparadigmchange} using the biomaRt R
package
\cite{Durinck.ea2009:Mappingidentifiersforintegration,Durinck.ea2005:BioMartandBioconductorpowerful}.
A total of \Sexpr{combined.fpkm[species!="homo sapiens" & mean_fpkm >=
  1,length(unique(human_name))]} 1:1 orthologous genes were expressed
in at least one species. Of these, \Sexpr{combined.fpkm[species!="homo
  sapiens" & mean_fpkm >= 10,length(unique(human_name))]} 1:1
orthologous genes with \ac{FPKM} $\ge 10$ were retained for future
analyses of gene expression evolution described below.

Significant shifts in expression levels between groups of species were
calculated using Student's t test on 1:1 orthologs using the mean of
each species to avoid large biases towards samples with $n \gg 1$ and
considered significant if the \ac{FDR} as estimated by the Benjamini
and Hochberg procedure \cite{Benjamini1995:bh_procedure} was
$\le 0.05$.

% We used the software \ac{OUCH} in the R statistical environment to
% investigate significant shifts in gene expression levels over
% time\cite{Butler.2004:phlogeneticcomparativeanalysis}. \ac{OUCH}
% institutes an \ac{OU} model of evolution, which allows for different
% optima for quantitative trait values along user-specified branches of
% a phylogenetic tree. The \ac{OU} model explicitly accounts for change
% in quantitative trait values by introducing two parameters into the
% model of stochastic Brownian motion of change in trait value along a
% phylogenetic tree through time. These parameters are $\theta$, an
% optimum trait value, and $\alpha$, the strength of selection
% (interpreted as “pull” toward the optimum). In addition, the \ac{OU}
% model can also integrate evolutionary change by assuming that each
% user-specified set of braches has its own separate \ac{OU} parameters
% \cite{Butler.2004:phlogeneticcomparativeanalysis}.
% 
% \DLA{The description of the OU process is terrible; it needs to be
%   completely redone}
% 
% Various \ac{OU} branch models were tested against a model in which no
% variation in optima ($\theta$) among branches was allowed. The
% following branches were specified in separate models to have a
% different optimum than all other branches: human, bonobo, mouse, dog,
% cow, elephant, opossum, human+bonobo, Euarchontoglires, and
% Boreoeutheria. Likelihood values were computed for each branch model
% and compared to an \ac{OU} model with one optimum using likelihood
% ratio tests with df=1. We filtered out genes with universally low
% expression levels, and included only genes in which expression in at
% least one species had an $\mathrm{FPKM}\ge10$ as described above. We
% corrected for multiple testing using the \ac{FDR} and using the
% Benjamini-Hochberg correction\cite{Benjamini1995:bh_procedure}. We
% used the time-calibrated phylogenetic tree of
% \cite{Meredith.ea2011:ImpactsofCretaceousTerrestrial} with all species
% not used in this analyses pruned.
% 
% We also calculated the enrichment of functional categories in the
% lists of genes which displayed a significant shift in expression on
% each branch using the same methodology as in the per-species gene
% lists \DLA{reference the tables once this analysis is complete}.

% \subsection{Correlation of Tissue Specific Index with Lineage Specific Genes}
% 
% To evaluate the extent of tissue specificity of each gene, we
% calculated the \ac{TSI} based on human RNASeq data in 3267 samples in
% 126 tissues the Roadmap Epigenomics
% \cite{Bernstein.ea2010:NIHRoadmapEpigenomicsMapping} and GTEx projects
% \cite{Gtexconsortium.2015:Thegenotypetissueexpression} as well
% additional placenta
% \cite{Hughes.ea2015:Evaluatingintra-andinter-individual} and uterus
% \cite{Chan.ea2014:Assessmentofmyometrialtranscriptome} samples from
% SRA. The \ac{TSI} measures the extent to which a specific gene has
% predominant expression in only one tissue
% \cite{Yanai.ea2005:Genome-widemidrangetranscriptionprofiles}. A gene
% expressed in only one tissue would have a \ac{TSI} of 1, whereas a
% gene expressed in all tissues would have a \ac{TSI} of 0.

% \subsection{Validation of Primate Expression using Spider Monkey Placenta Transcriptome}
% 
% To verify expression levels in human and bonobo, we also generated
% RNA-Seq data for the placenta transcriptome of the black-headed spider
% monkey (\textit{Ateles fusciceps}) using similar protocols discussed above.
% Ateles is outside the clade, which includes hominids and can be used
% to polarize expression values of highly expressed genes in Homo and
% Pan. We sampled placental tissue from an Ateles placenta expelled from
% the mother shortly after birth (see Supplemental Table X). We used
% similar procedures discussed above to extract and sequence this
% sample. [DEREK—I’m unsure how to proceed on describing this so as to
% not step on Natalie’s toes]
% 
% \DLA{Complete discussion of the ateles placenta}

\subsection{Sequence Deposition and Code}

All of the code used to produce every analysis presented in this paper
is available at
\url{https://github.com/uiuc-cgm/placenta-viviparous.git}. All of the
sequences used are deposited in SRA, and can be obtained under the
series
\href{http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE79121}{GSE79121}.

>>>>

All 1:1 orthologs present among \opossum, \textit{Pan troglodytes},
\cow, \dog, \elephant, \mouse, and \human were determined using
Ensembl v.80 gene trees via Biomart
\cite{Kasprzyk2011:BioMartdrivingparadigmchange} using the biomaRt R
package
\cite{Durinck.ea2009:Mappingidentifiersforintegration,Durinck.ea2005:BioMartandBioconductorpowerful}.
A total of \Sexpr{combined.fpkm[species!="homo sapiens" & mean_fpkm >=
  1,length(unique(human_name))]} 1:1 orthologous genes were expressed
in at least one species. Of these, \Sexpr{combined.fpkm[species!="homo
  sapiens" & mean_fpkm >= 100,length(unique(human_name))]} 1:1
orthologous genes with \ac{FPKM} $\ge 100$ were retained for future
analyses of gene expression evolution described below.

Significant shifts in expression levels between groups of species were
calculated using Student's t test on 1:1 orthologs using the mean of
each species to avoid large biases towards samples with $n \gg 1$ and
considered significant if the \ac{FDR} as estimated by the Benjamini
and Hochberg procedure \cite{Benjamini1995:bh_procedure} was
$\le 0.05$. For \human and \mouse lineage specific calculations, a
type-III anova was calculated using
lme4~\cite{Bates.ea2015:FittingLinearMixed-EffectsModels} with each
species having separate random effects.

\subsection{Sequences, FPKMs, and Code}

All of the code used to produce every analysis presented in this paper
is available at
\url{https://github.com/uiuc-cgm/placenta-viviparous.git}. All of the
sequences used are deposited in SRA, and new sequences for this work
can be obtained under the series
\href{http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE79121}{GSE79121}
(Supplemental \cref{tab:srr_numbers}). The mean FPKM for gene in each
species with ortholog annotations are located in the supplemental data
files \Sexpr{english.join(paste0(gsub(" ","_",
  star.logs[,unique(species)]),"_mean_fpkm.txt"))}.
